Below-ground sink limitation alters growth and carbon balance of Eucalyptus seedlings

### 2015-09-23 manuscript/manuscript_pve.Rmd
edits MT

Interpreting limitations to plant growth requires understanding of the balance between carbon (C) source and sink activity. This study used manipulations of soil volume to test how growth is coupled to physiology, allocation, and sink activity in *Eucalyptus tereticornis* seedlings. We grew individual seedlings in a large range of container sizes and planted containers flush to the soil alongside naturally sown ('free') seedlings. Reduced soil volume was expected to induce rapid negative effects on growth and physiology compared to free seedlings. It was hypothesized that the soil volume effect would be largest in the smallest containers, negatively impacting mass partitioning belowground. An accumulation of leaf non-structural carbohydrates, resulting from reduced belowground sink strength, was expected to correlate to reductions in photosynthetic capacity. We observed a negative effect of contatiner volume on aboveground growth soon after the experiment started. Although growth was consistently different across soil volumes, dry mass partitioning to leaves, stems and roots was unchanged after 120 days. Photosynthetic capacity was significantly reduced in containers, and was related to both leaf nitrogen content and starch accumulation. We developed a seedling growth model that utilized leaf photosynthesis rates (A) to allocate daily C uptake towards mass growth of stems, leaves and roots. We then asked whether the observed reductions in A explained the observed differences in seedling biomass. We found that although belowground sink limitation resulted in down regulation of A, these reductions were not large enough to explain observed growth responses. Thus, as A and growth were not tightly coordinated, an unaccounted for pool of non-biomass C resulted in seedlings with soil volume restriction. These results highlight the need to further utilize mass balance approaches when evaluating plant C allocation and confirms that A and growth are not always directly related.

@@

## Gas exchange and photosynthetic parameters
A~sat~ and A~max~ were both significantly higher in the largest container volume and the free seedling at the first measurement campaign (both P < 0.001). Across all measurement campaigns A~sat~ (Figure `r asatfig`) and A~max~ (Table 2.) were consistently higher in free seedlings than in containers (26 % and 29 %, respectively). The interaction between photosynthetic capacity, leaf starch, and leaf N on a mass basis was marginally significant (P = 0.058) but A~max~ was highly correlated to both leaf N content and leaf starch (both P < 0.001). Across all measurement campaign A~max~ was higher when foliar N was also higher, usually associated with low levels of leaf starch (Figure `r achemfig`a). A~max~  was also lower when leaf starch was high as higher leaf N often did not coincide with high leaf starch (Figure `r achemfig`b). Overall, A~max~ was significantly correlated with final harvest biomass across all seedlings (P < 0.001)

\
Both J~max~ and Vc~max~ were significantly higher in free seedlings (30 % and 26 %, respectively) with little variation between soil volume treatments (Table 2). Leaf dark respiration rates were not significantly different across soil volumes (Table 2). The g~1~ parameter, generated for each seedling from the Medlyn et al [-@medlyn2011reconciling] optimal stomatal conductance model, was lowest in the free seedling and was marginally different across soil volume treatments (Table 2).

\
Neither $\Psi$~pd~ nor $\Psi$~l~ were different across treatments, with mean values of -0.27 and -1.2 MPa across all seedlings, respectively. Although g~s~ in free seedlings was generally higher than those in containers (Table 2. P = 0.001), the mean rates for all seedlings were high at 0.37 mol m^-2^ s^-1^ and did not decline significantly across the experiment duration. Additionally, leaf $\delta$^13^C at final harvest was not different across treatments (Table 1). Combined these indices provide strong evidence that water stress was not apparent on these seedlings throughout the experiment. Soil N % at harvest was not different across soil volumes ($\bar{x}$ = 04.5 %) and decreased approximately 3 % across all containers over the experiment duration. This indicates that nutrient leaching from free seedlings or from draining of containers following natural rainfall events did not differ between treatments.

## Modelling seedling biomass
Model M~0~, was optimized so that the free seedling would fit as well as possible and then used to predicted biomass for seedlings in containers. M~0~ was able to converge on an optimum LMF (21.6 %) which predicted the mean harvest total biomass of free seedlings within 1.2 %. Using this optimized LMF, the total biomass of modeled seedlings with soil volume restriction (M~1~) was on average 23&pm;2.4 g C more than measured seedlings when comparing against predicted total net leaf C gain (Figure `r massmodelfig`a). Thus, seedling C mass was overestimated by an average of 50&pm;8.7 % in modeled seedlings across the reductions in daily C assimilation per treatment (Figure `r massmodelfig`b). As a result, the reductions in leaf A were not large enough to explain the reduction in observed seedling biomass across the  container size treatments.

\
Next, we performed a series of model simulations to test possible C allocation scenarios to account for this excess pool of C. Testing the sensitivity of the model to treatment-specific absolute LMF from the final harvest (M~2~) improved model predictions but still overestimated seedling total C by 32&pm;11.1 % (Figure `r massmodelfig2`a). Using harvested LMF, however, does not capture the increase in senescence of seedlings in small containers (Figure `r allomfig`c). Thus, the 7-61 % reduction in harvested LMF from large to small soil volumes compared to the optimized LMF (M~0~) represents an underestimation of realized leaf C allocation. Increases of 50 % in non-leaf tissue respiration (M~3~) improved biomass estimates slightly but overestimated mass C by an average of 46&pm;9.3 % in seedlings with soil volume restriction (Figure `r massmodelfig2`b). With M~3~, non-tissue respiration rates would need to be increased by ca. 250% in order for mass balance to be achieved.

# Discussion
This study utilized a simple but novel field design to manipulate belowground sink limitation and physically restrict *Eucalyptus tereticornis* seedling production. We addressed questions regarding the coordination of A and growth by complementing empirical results with a carbon balance model. We found that reductions in leaf A across container sizes were not sufficient to explain observed reductions in total plant biomass production. We thus encourage the utilization of mass balance approaches and provide direction for future studies when testing factors that control plant growth under environmental change.

## Reductions in growth and physiology under sink limitation
Soon after seedlings became established both height and diameter growth were negatively affected by decreasing soil volume. This led to the large reductions in biomass in small containers, compared to freely rooted seedlings. These growth reductions were expected, as the impedance of root growth can cause reductions in overall plant growth and activity [@mcconnaughay1991physical; @young1997mechanical]. It has been shown that roots undergoing difficult conditions may send inhibitory signals to the shoots that affect g~s~, cell expansion, cell division and the rate of leaf appearance [@passioura2002soil].  Here, this was evident in a large divergence in leaf area between seedlings in containers and free seedlings through time, with the eventual cessation of new leaf growth in seedlings in small containers.

\
Decreases in A~sat~ occurred at the same time as reductions in allometric growth parameters of seedlings in containers. This initially suggests a strong link between growth and an apparent down regulation of A. However, there are several possible mechanisms that can explain reduced A in small pots including nutrient content, water or reduced sink strength [@poorter2012pot].  It was therefore necessary to examine each of these factors to distinguish if the induced belowground sink limitation actually triggered photosynthetic down regulation.

\
With high rates of g~s~, non-limiting leaf water potential and consistent leaf $\delta$^13^C across soil volume treatments there was little evidence of water stress causing the reduction in A. This finding is consistent with other container size studies without drought treatments.  For example, reduced A~max~ in cotton seedlings grown at elevated CO~2~ was attributed to sink-limited feedback inhibition from inadequate rooting volume, not decreased g~s~ [@thomas1991root]. Additionally, severe reductions in A in coffee plants was not attributed to impacts of container size on leaf water potentials or g~s~ [@ronchi2006growth]. It is likely that reductions in A of these well watered seedlings was instead the result of limiting soil nutrients or barrier sensing on belowground sink strength.

\
Although the soil N pool declined evenly across all treatments, leaf N was lowest in the smallest containers, suggesting sink limitation was the greatest in these containers. This makes sense as small containers may result in a reduction of total N uptake, either from physical root restriction or decreased supply, which will affect growth, Rubisco limitation, sugar metabolism, and carbohydrate partitioning between source and sink tissues [@stitt1991rising; @hermans2006plants]. Mycorrhizal colonization could also have been affected in containers. Whether unrestricted mycorrhizal recruitment facilitated the increase in leaf N uptake in free seedlings is unknown. Regardless of whether sink limitation occurred from root or mycorrhizal restriction, the stark contrast in leaf N between free seedlings and seedlings in containers was significantly correlated to reductions in A.

\
As both rooting space and resources were finite in containers, the inability of seedlings to increase the capacity of the belowground C sink resulted in the buildup of C assimilate in leaves. The feedback inhibition of A from starch accumulation has been proposed, yet it is still not known whether there is a starch threshold that triggers the down-regulation process (Nebauer et al. 2011).  Here, declines in A~max~ were correlated with higher starch content throughout the experiment. This agrees with a study on a deciduous conifer by Equiza et al. (2006) where photosynthetic downregulation from reduced sink strength was more correlated with starch than sugar content. As starch content in the smallest containers was nearly double that of free seedlings in our study, this suggests the response of A to sink inhibition was regulated by this accumulation, as hypothesized.

## Biomass partitioning under sink limitation
As biomass partitioning is likely controlled by the source and sink strength of all organs [@poorter2012biomass] it was important to determine which tissue components were most affected by the container size treatments. It was necessary to distinguish if growth was affected beyond ontogenetic constraints, by correcting for size, as biomass distribution is strongly size-dependent [@gould1966allometry; @lleonart2000removing]. In this study, there was no significant difference in root, leaf, or stem biomass partitioning with reduced soil volume compared to free seedlings, outside of ontogenetic drift. This is a surprising result as shifts in allocation have been noted specifically for nutrient limitation  [@mcconnaughay1999biomass, and references therein]. Surprisingly, fine root to leaf mass ratio was conserved across all treatments suggesting a functional partitioning response to optimize resource gain did not occur.

@@

\
**Figure 5**. Photosynthetic capacity, on a leaf mass basis, as a function of accumulation of leaf starch (a) and leaf nitrogen content without TNC (b).  Colors represent bins levels (n = 5) of both leaf starch and nitrogen grouped from low to high .  Lines represents predictions, for each bin level, from the linear mixed effects model equation of A~max~ as a function of starch and nitrogen. The marginal R^2^ (fixed effects only) was 0.37 and the conditional R^2^ (fixed and random effects) was 0.48 for the complete model.

>>>>

Interpreting limitations to plant growth requires understanding of the balance between carbon (C) source and sink activity. This study used manipulations of soil volume to test how growth is coupled to physiology, carbon allocation, and sink activity in *Eucalyptus tereticornis* seedlings. We grew individual seedlings in a large range of container sizes and planted containers flush to the soil alongside naturally sown ('free') seedlings. We developed a seedling growth model that utilized leaf photosynthesis rates (A) to allocate daily C uptake towards mass growth of stems, leaves and roots. Reduced soil volume was expected to induce rapid negative effects on growth and physiology compared to free seedlings. It was hypothesized that the soil volume effect would be largest in the smallest containers, negatively impacting mass partitioning belowground. An accumulation of leaf non-structural carbohydrates, resulting from reduced belowground sink strength, was expected to correlate to reductions in photosynthetic capacity. We observed a negative effect of contatiner volume on aboveground growth soon after the experiment started. Although growth was consistently different across soil volumes, dry mass partitioning to leaves, stems and roots was unchanged after 120 days. Photosynthetic capacity was significantly reduced in containers compared to free seedlings, and was related to both leaf nitrogen content and starch accumulation. We then asked whether the observed reductions in A explained the observed differences in seedling biomass. We found that although belowground sink limitation resulted in down regulation of A, these reductions were not large enough to explain observed growth responses. Thus, as A and growth were not tightly coordinated, an unaccounted for pool of non-biomass C resulted in seedlings with soil volume restriction. These results highlight the need to further utilize mass balance approaches when evaluating plant C allocation and confirms that A and growth are not always directly related.

@@

## Gas exchange and photosynthetic parameters
A~sat~ and A~max~ were each significantly higher in the largest container volume and the free seedling treatment compared to the smaller containers at the first measurement campaign (both P < 0.001). Across all measurement campaigns mean A~sat~ (Figure `r asatfig`) and A~max~ (Table 2) were consistently higher in free seedlings than in containers (26 % and 29 %, respectively). The relationships between photosynthetic capacity, leaf starch, and leaf N on a mass basis was marginally significant (P = 0.058) but A~max~ on a mass basis was highly correlated to both leaf N content and leaf starch (both P < 0.001). Across all measurement campaign A~max~ was higher when foliar N was also higher, usually associated with low levels of leaf starch (Figure `r achemfig`a). A~max~  was also lower when leaf starch was high as higher leaf N often did not coincide with high leaf starch (Figure `r achemfig`b). Overall, A~max~ was significantly correlated with final harvest biomass across all seedlings (P < 0.001)

\
Both J~max~ and Vc~max~ were significantly higher in free seedlings (30 % and 26 %, respectively) than container-grown seedlings with little variation between container volume treatments (Table 2). Leaf dark respiration rates were not significantly different across soil volumes (Table 2). The g~1~ parameter, generated for each seedling from the Medlyn et al [-@medlyn2011reconciling] optimal stomatal conductance model, was lowest in the free seedling treatment and was marginally different across soil volume treatments (Table 2).

\
Neither $\Psi$~pd~ nor $\Psi$~l~ were different across treatments, with mean values of -0.27 and -1.2 MPa across all seedlings, respectively. Although g~s~ in free seedlings was generally higher than those in containers (Table 2. P < 0.001), the mean rates for all seedlings were high at 0.37 mol m^-2^ s^-1^ and did not change throughout the course of the experiment. Additionally, leaf $\delta$^13^C at final harvest was not different across treatments (Table 1). Combined these indices provide strong evidence that water stress was not apparent on these seedlings throughout the experiment. Soil N % at harvest was not different across soil volumes ($\bar{x}$ = 4.5 %) and decreased approximately 3 % across all containers over the experiment duration. This indicates that nutrient leaching from free seedlings or from draining of containers following natural rainfall events did not differ between treatments.

## Modelling seedling biomass
Model M~0~, was optimized so that the free seedling would fit as well as possible and then used to predicted biomass for seedlings in containers. M~0~ was able to converge on an optimum LMF (21.6 %) which predicted the mean harvest total biomass of free seedlings within 1.2 %. Using this optimized LMF, the total biomass of modeled seedlings with soil volume restriction (M~1~) was on average 23&pm;2.4 g C more than measured seedlings when comparing against predicted total net leaf C gain (Figure `r massmodelfig`a). Thus, seedling C mass was overestimated by an average of 50&pm;8.7 % in modeled seedlings across the soil volume treatments (Figure `r massmodelfig`b). As a result, the observed reductions in leaf A with decreasing soil volume when integrated across the 120 day experiment were not large enough to explain the reduction in observed seedling biomass across the  container size treatments.

\
Next, we performed a series of model simulations to test possible C allocation scenarios to account for this excess pool of C. Testing the sensitivity of the model to treatment-specific LMF from the final harvest (M~2~) improved model predictions of seedling C mass but still overestimated seedling total C by 32&pm;11.1 % (Figure `r massmodelfig2`a). Using harvest values of LMF, however, does not capture the increase in senescence of seedlings in small containers (Figure `r allomfig`c). Thus, the 7-61 % reduction in harvest avlues of LMF from large to small soil volumes compared to the optimized LMF (M~0~) represents an underestimation of realized leaf C allocation. Increases of 50 % in non-leaf tissue respiration (M~3~) improved biomass estimates slightly but overestimated mass C by an average of 46&pm;9.3 % in seedlings with soil volume restriction (Figure `r massmodelfig2`b). With M~3~, non-tissue respiration rates would need to be increased by ca. 250% in order for mass balance to be achieved.

# Discussion
This study utilized a simple but novel field design to manipulate belowground sink limitation and physically restrict *Eucalyptus tereticornis* seedling biomass production. We addressed questions regarding the coordination of A and growth by complementing empirical results with a carbon balance model. We found that reductions in leaf A across container sizes were not sufficient to explain observed reductions in total plant biomass production. We thus encourage the utilization of mass balance approaches and provide direction for future studies when testing factors that control plant growth under environmental change.

## Reductions in growth and physiology under sink limitation
Soon after seedlings became established both height and diameter growth were negatively affected by decreasing soil volume. This led to the large reductions in biomass in small containers, compared to freely rooted seedlings. These growth reductions were expected, as the impedance of root growth can cause reductions in overall plant growth and activity [@mcconnaughay1991physical; @young1997mechanical]. It has been shown that roots subjected to environmental stress may send inhibitory signals to the shoots that affect g~s~, cell expansion, cell division and the rate of leaf appearance [@passioura2002soil].  Here, this was evident in a large divergence in leaf area between seedlings in containers and free seedlings through time, with the eventual cessation of new leaf growth in seedlings in small containers.

\
Decreases in A~sat~ occurred at the same time as reductions in allometric growth parameters of seedlings in containers. This initially suggests a strong link between growth and an apparent down regulation of A. However, there are several possible mechanisms that can explain reduced A in small pots including nutrient content, water or reduced sink strength [@poorter2012pot].  It was therefore necessary to examine each of these factors to determine if the induced belowground sink limitation actually triggered photosynthetic down regulation.

\
With high rates of g~s~, non-limiting leaf water potential and consistent leaf $\delta$^13^C across soil volume treatments there was little evidence that water stress caused the reduction in A. This finding is consistent with other container size studies without drought treatments.  For example, reduced A~max~ in cotton seedlings grown at elevated CO~2~ was attributed to sink-limited feedback inhibition from inadequate rooting volume, not decreased g~s~ [@thomas1991root]. Additionally, severe reductions in A in coffee plants were not attributed to impacts of container size on leaf water potentials or g~s~ [@ronchi2006growth]. It is likely that reductions in A of well-watered seedlings oberved in our study of *E.tereticornis* seedlings was instead the result of limiting soil nutrients or barrier sensing on belowground sink strength.

\
Although the soil N % declined evenly across all treatments, leaf N was lowest in the smallest containers, suggesting sink limitation was the greatest in these containers. This makes sense as small containers may result in a reduction of total N uptake, either from physical root restriction or decreased supply, which will affect growth, Rubisco limitation, sugar metabolism, and carbohydrate partitioning between source and sink tissues [@stitt1991rising; @hermans2006plants]. Mycorrhizal colonization could also have been affected in containers. Whether unrestricted mycorrhizal recruitment facilitated the increase in leaf N uptake in free seedlings is unknown. Regardless of whether sink limitation occurred from root or mycorrhizal restriction, the stark contrast in leaf N between free seedlings and seedlings in containers was significantly correlated to reductions in A.

\
As both rooting space and resources were finite in containers, the inability of seedlings to maintain the capacity of the belowground C sink resulted in the buildup of C assimilate in leaves. The feedback inhibition of A from starch accumulation has been proposed, yet it is still not known whether there is a starch threshold that triggers the down-regulation process (Nebauer et al. 2011).  Here, declines in A~max~ were correlated with higher starch content throughout the experiment. This agrees with a study on a deciduous conifer by Equiza et al. (2006) where photosynthetic downregulation from reduced sink strength was correlated with starch content. As starch content in leaves of plants grown in the smallest containers was nearly double that of free seedlings in our study, this suggests the response of A to sink inhibition was regulated by this accumulation, as hypothesized.

## Biomass partitioning under sink limitation
As biomass partitioning is likely controlled by the source and sink strength of all organs [@poorter2012biomass], it was important to determine which tissue components were most affected by the container size treatments. It was necessary to distinguish if growth was affected beyond ontogenetic constraints, by correcting for size, as biomass distribution is strongly size-dependent [@gould1966allometry; @lleonart2000removing]. In this study, there was no significant difference in root, leaf, or stem biomass partitioning with reduced soil volume compared to free seedlings, outside of ontogenetic drift. This is a surprising result as shifts in allocation have been noted specifically for nutrient limitation  [@mcconnaughay1999biomass, and references therein]. Surprisingly, fine root to leaf mass ratio was conserved across all treatments suggesting a functional partitioning response to optimize resource gain did not occur.

@@

\
**Figure 5**. Photosynthetic capacity, on a leaf mass basis, as a function of accumulation of leaf starch (a) and leaf nitrogen content without TNC (b).  Colors represent bins levels (n = 5) of both leaf starch and nitrogen grouped from low to high .  Lines represents predictions, for each bin level, from the linear mixed effects model equation of A~mass~ as a function of starch and nitrogen. The marginal R^2^ (fixed effects only) was 0.37 and the conditional R^2^ (fixed and random effects) was 0.48 for the complete model.
